---
title: "session08 - homework"
format: html
editor: visual
date: 2025-07-17
author: Timo Roettger
execute:
  error: false
  warning: false
  message: false
  cache: true
bibliography: ../resources/bibliography.bib
---

# Preamble: Loading packages and configuration

```{r}
#| label: data_and_libraries
#| echo: false

# just run this code chunk
# function to ignoring the setting of the relative path below when knitting
run_if_not_knitting <- function(expr) {
  if (!isTRUE(getOption("knitr.in.progress"))) {
    eval(expr)
  }
}

# nifty code using the pacman package
# it checks if the packages specified below are installed, if not, they will be installed, if yes, they will be loaded
if (!require("pacman")) install.packages("pacman")
pacman::p_load(rstudioapi, tidyverse, tidybayes, brms, bayesplot, bayestestR, modelr)

# set the current working directory to the one where this file is
run_if_not_knitting(current_working_dir <- dirname(rstudioapi::getActiveDocumentContext()$path))
run_if_not_knitting(setwd(current_working_dir))

# options to increase efficiency of brms models (optional)
#options(brms.backend = "cmdstanr")
options(mc.cores = parallel::detectCores())

```

# Exercise

We will be re-analyzing the data from @winter2013making.

```{r}
#| label: preprocessing_sim

# Load our datasets
sim <- read_csv("../data/linguistic_similarity.csv") |> 
  mutate(Sem = as.factor(Sem),
         Phon = as.factor(Phon))
sim

```

We will now retrace the original article's complete analysis. In their original analysis, they were interested in whether (a) semantic similarity between cities, (b) phonological similarity between cities, and (c) their interaction affects how close/far apart participants draw the cities.

## (a) Specify Priors

Specify the model formula and appropriate weakly informative priors for all relevant predictors. Contrast-code the predictors.

```{r}
#| label: model_formula_sim

contrasts(sim$Sem) <- contr.sum(2)
contrasts(sim$Phon) <- contr.sum(2)

# For both, Similar = -1 and Diff = 1

# define formula
main_formula <- bf(Distance ~ Sem * Phon)

# get priors
get_prior(main_formula, sim)

# set priors
# let's assume every point on the paper (~250mm^2 area) is equally likely, and there are no condition effects. SD for all effects can be 100 (~half the size of the paper). So anything could happen, as far as I know.
my_priors <- c(prior(uniform(0, 250), class = 'Intercept'),
               prior(normal(0, 100), class = 'b', coef = 'Phon1'),
               prior(normal(0, 100), class = 'b', coef = 'Sem1'),
               prior(normal(0, 100), class = 'b', coef = 'Sem1:Phon1'))

```

## (b) Run the appropriate model for the research question

```{r}
#| label: model_sim

model_posterior <- brm(main_formula,
                  family = gaussian(),
                  prior = my_priors,
                  data = sim,
                  chains = 4,
                  cores = 4, 
                  iter = 2000,
                  warmup = 1000,
                  thin = 1,
                  seed = 42)


```

## (c) Posterior predicive check

Check the overall model fit with posterior predictive checks. If the model does not fit the data well, fix the problem (check `06_homework_solutions.qmd` for inspiration).

```{r}
#| label: model_fit

pp_check(model_posterior, ndraws = 100) # seems to match the shape of the data rather well in terms of distribution of predicted distances (DVs). As well as any of our past models, at least. The peak is lower and the right side is higher and doesn't catch the waviness of the data. 
```

## (d) Assess sampling diagnostics

Check that chains are reasonably mixed and that sampling yielded a sufficient number of effective sample size. If there are issues, address them accordingly.

```{r}
#| label: sampling_diagnostics

model_posterior # ~5000 Bulk_ESS and ~3000 Tail_ESS for all params--looks good

mcmc_plot(model_posterior, type = 'trace') # good caterpillars

```

## (e) Interpret the results

Now interpret the model relative to the research question and describe uncertainty, visualize posteriors, calculate posterior probabilities of direction, and calculate Bayes Factors for evidence for and against the posterior coefficients being point zero.

(Note: When you want to extract interaction terms like `b_Sem1:Phon1`, you have to put single quotation marks around them because R struggles with interpreting the colon)

```{r}
#| label: interpret_sim

# get posteriors
get_variables(model_posterior)
interaction_name <- "b_Sem1:Phon1" # spread_draws doesn't like the colon
posteriors <- 
  model_posterior |> 
  spread_draws(b_Intercept, b_Sem1, b_Phon1, !!sym(interaction_name)) |> 
  select(b_Intercept, b_Sem1, b_Phon1, !!sym(interaction_name))

# Uncertainty
model_posterior # Everything has an estimated error of ~2, which is fine for the intercept (makes estimates range 73-81), but worse for Sem1 (estimate ~5, range 0.9-9), and terrible for Phon1 and Sem1:Phon1, which both have negative estimates (-2.9 and -1.12 respectively) whose distributions cross 0.

# Visualize posteriors
model_posterior |> 
  spread_draws(b_Intercept, b_Sem1, b_Phon1, !!sym(interaction_name)) |> 
  ggplot(aes(alpha = 0.4)) +
   stat_halfeye(aes(x = b_Intercept, fill = 'Intercept')) +
   stat_halfeye(aes(x = b_Sem1, fill = 'Sem')) +
   stat_halfeye(aes(x = b_Phon1, fill = 'Phon')) +
   stat_halfeye(aes(x = !!sym(interaction_name), fill = interaction_name)) +
   geom_vline(xintercept = 0, lty = 'dashed') +
   labs(x = "Distance (mm)",
        y = "Density") +
   theme_minimal()
# All touch/cross 0 except for Intercept. This sets off alarm bells but we'll want to see PD to quantify what's goin on.

# PD
p_direction(posteriors) # as hinted at with the uncertainty and visual inspection, the Sem1 and Phon1 parameters have lower probability of direction (99% and 92%), since some of their samples cross over 0. And then the interaction effect is especially problematic (71%).

# Bayes Factors relative to 0
model_prior <- brm(main_formula,
                  family = gaussian(),
                  prior = my_priors,
                  sample_prior = "only",
                  data = sim,
                  chains = 4,
                  cores = 4, 
                  iter = 2000,
                  warmup = 1000,
                  thin = 1,
                  seed = 42)

priors <- model_prior |> 
  spread_draws(b_Intercept, b_Sem1, b_Phon1, !!sym(interaction_name)) |> 
  select(b_Intercept, b_Sem1, b_Phon1, !!sym(interaction_name))

bayesfactor_parameters(posteriors, priors, direction = "two-sided", null = 0)
# I got a warning about the Bayes factor having 2701 divergent transitions after warmup.
# Intercept: 2.29e+39 Extremely strong evidence against the idea that all points on the map are equally likely. They are far more likely to be around ~77mm away from each other.
# Sem1: 0.4           Anecdotal evidence in favor of the null. It's really too close to call, though, so we might not be able to trust this parameter's inferences at all.
# Phon1: 0.058        Strong evidence in favor of the null, implying that there is no meaningful effect of Phonological similarity on distance.
# Sem1:Phon1: 0.023   Quite strong evidence in favor of our priors/null hypothesis thingy, which was that there is no meaningful interaction effect.

```

## (g) Write up your entire analysis:

I will reiterate here the three research interests: whether (a) semantic similarity between cities, (b) phonological similarity between cities, and (c) their interaction affects how close/far apart participants draw the cities. Each of these was assigned a parameter in a Bayesian model, with a uniform prior for distance between 0 and 250, and effect priors centered at 0 with a standard deviation of 100.

The posterior distributions of the model indicate that the effect of semantic similarity is an estimated increase in distance by an average of 5.03 mm (95% Credible Interval: 0.91 to 9.05). For phonological similarity, the effect is an estimated decrease, such that the average distance changes by 2.9 mm (95% Credible Interval: -6.75 to 1.03). For the interaction effect (i.e., when cities are both semantically and phonologically similar), an additional parameter would shift the estimated distance by an average of -1.12 mm (95% Credible Interval: -5.18 to 3.02). For conditions of semantic or phonological difference, as well as their interaction, the opposite effect of the appropriate parameter would apply, since the data was contrast-coded.

I interpret these uncertainty intervals as problematic, as many of the samples indicate an effect that is either extremely small or sometimes even contradictory to the other samples (that is, of an inverse signage). This issue is also evident from visual inspection of the posterior distributions. When calculating the probability of direction (also known as Maximum Probability of Effect), the samples for the Semantic and Phonological parameters had imperfect, but still quite high, probabilities of direction (99% and 92%, respectively), since only some of their samples cross over 0. The samples for the interaction effect was especially incoherent (71%). This casts doubt on the capability of the model to address research interest (c). \[is this true? or is it simply evidence of an effect of 0\]

Subsequently, I calculated the Savage-Dickey density ratio [@DickeyLientz1970-The-Weighted-Li; @WagenmakersLodewyckx2010-Bayesian-hypoth; @roettgermeaningful], relating the evidence for the model based solely on priors to the evidence for the model based on both priors and likelihood [e.g., @WetzelsGrasman2010-An-encompassing]. My analysis yielded a Bayes Factor of 0.4 for the semantic parameter, indicating the model produced only anecdotal evidence in favor of the null hypothesis. The Bayes Factor for the phonological parameter was 0.058, indicating strong evidence in favor of the null hypothesis. The Bayes Factor for the interaction effect was 0.023, indicating very strong evidence in favor of the null hypothesis.

Taken together, my analysis indicates that there are no meaningful effects of (a) semantic similarity, (b) phonological similarity, or (c) an interaction of the two conditions on how far apart participants draw cities.

## OPTIONAL (h) compare your interpretation to the original findings

YOUR COMPARISON HERE

# References

<!-- References will auto-populate in the refs div below -->

::: {#refs}
:::
