---
title: "session11 - homework"
format: html
editor: visual
date: 2025-07-17
author: Timo Roettger
execute:
  error: false
  warning: false
  message: false
  cache: true
bibliography: ../resources/bibliography.bib
---

# Preamble: Loading packages and configuration

```{r}
#| label: data_and_libraries
#| echo: false

# just run this code chunk
# function to ignoring the setting of the relative path below when knitting
run_if_not_knitting <- function(expr) {
  if (!isTRUE(getOption("knitr.in.progress"))) {
    eval(expr)
  }
}

# nifty code using the pacman package
# it checks if the packages specified below are installed, if not, they will be installed, if yes, they will be loaded
if (!require("pacman")) install.packages("pacman")
pacman::p_load(rstudioapi, tidyverse, tidybayes, brms, ggdist, bayestestR, modelr)

# set the current working directory to the one where this file is
#run_if_not_knitting(current_working_dir <- dirname(rstudioapi::getActiveDocumentContext()$path))
#run_if_not_knitting(setwd(current_working_dir))

# options to increase efficiency of brms models (optional)
#options(brms.backend = "cmdstanr")
options(mc.cores = parallel::detectCores())

```

# Introduction

We will be analyzing the data from @roettger2014assessing.

```{r}
#| label: preprocessing

# Load our datasets
incNeutr <- read_csv("../../data/E1_production.csv") |> 
  # we will focus only on the relevant columns
  select(subject, voicing, item_pair, usable, vowel_dur) |> 
  # and only on rows the authors deemed usable
  filter(usable == "usable") |> 
  select(!usable)

```

The variables:

`subject`: unique identifier for participants (n = 16)

`voicing`: underlying voicing status of stop (`voiced` vs. `voiceless`)

`item_pair`: unique identifier for pseudo word minimal pairs (n = 24)

`vowel_dur`: vowel duration in ms

In this study, we recorded people producing words such as Rad and Rat in German and were interested to see whether these words phonetically differ. The reason why this is interesting is that, according to German phonology, these words *should* be identical, i.e. he voicing distinction between /d/ and /t/ should have been *neutralized.* However, previous literature indicates that this neutralization is phonetically *incomplete.* For example, the vowel preceding a (de)voiced stop such /d/ is slightly longer preceding a voiceless stop such as /t/.

The researchers hypothesize that the vowel duration is meaningfully longer preceding voiced stops than preceding voiceless stops. They model this relationship using a linear mixed effects model that accounts for both speaker and word variation as well as by-speaker and by-word variation for the critical predictor voicing.

```{r}
#| label: model_formula

# store model formula
mdl_formula = bf(vowel_dur ~ voicing +
                   # random intercepts and slopes for subject and item_pair
                   (1 + voicing | subject) +
                   (1 + voicing | item_pair))

```

```{r}
#| label: spec-priors

# some weakly informative priors 
my_priors <- c(prior(normal(150,50), class = 'Intercept'),
               # weakly informative prior for the voicing difference (very diffuse)
               prior(normal(0, 30), class = 'b', coef = 'voicingvoiceless'),
               # weakly informative prior for all sources of variance (very diffuse)
               prior(cauchy(0,100), class = 'sd'),
               # weakly informative prior for residual variance (very diffuse)
               prior(cauchy(0,100), class = 'sigma'),
               # correlations
               prior(lkj(1), class = 'cor'))

```

```{r}
#| label: run-model

# run model
xmdl <- brm(mdl_formula,
            family = gaussian(),
            prior = my_priors,
            file = "../../models/in_sesoi",
            data = incNeutr,
            seed = 42)

# I also store the prior-only model 
xmdl_prior <- brm(mdl_formula,
            family = gaussian(),
            prior = my_priors,
            sample_prior = "only",
            file = "../../models/in_sesoi_priors",
            data = incNeutr,
            seed = 42)

```

# Exercises

## (a) Plot the posterior distribution

Explore the posterior distribution for the difference between voiced and voiceless stops. Interpret it.

```{r}

xmdl %>% spread_draws(b_voicingvoiceless) %>% 
  ggplot(aes(x = b_voicingvoiceless)) +
  stat_halfeye() +
  labs(x = "\nEst. difference from (de)voiced to voiceless",
       y = "Density\n") +
  theme_minimal()

```

## (b) Define a SESOI

Define a SESOI and justify it. Can you find an applied justification for a SESOI. Or a theoretical justification? Are there justifications grounded in measurement? What about justifications based on previous research? Define it and formalize it as a Region of Practical Equivalence as we have done before. Use the `bayesfactor_rope()` function to calculate the Bayes factor for and against the posteriors being inside the ROPE.

**Lines of SESOI inquiry:** theoretical, experimental, methodological evidence for SESOIs.

**Theoretical:** Is smallest of interest = the smallest perceptible? Even an imperceptible difference would still have something to say about acoustic or articulatory phonetics, but would not be grounds for a maximalist claim that both voiced and devoiced phonemes retain a conscious distinction as evinced by perceived minimal pairs. Weaker claims could still tie into, for example, the phenomenon of vowels lengthening even before non-neutralized voiced stops. What's more is that unconscious differences could still be sociolinguistically relevant (can they propogate unconsciously?) or useful for language technology. SO I WOULD SAY that importance/meaningfulness is actually orthogonal to just-noticeability.

**Experimental:** Previous works cited in the OG 2014 paper claim 1.2-6.2 ms as the effect size of IN (Port and Crawford, 1989). So setting a SESOI of even 1ms might be pushing it, if we want our analysis to be relevant to theirs. As described above, we don't care about a just-noticeable difference, but for what it's worth I found literature attesting as low as 5ms and as high as 33-100ms for vowel JNDs. And then the vast majority, according an LLM's summary of the literature (sorry), point to between 5 and 15 ms as plausible IN duration differences.

**Methodological:** In this paper's methods section, I didn't see anything about the measurements, task, or items that indicated a methodological lower bound. But we could maybe say 1ms, since that's the smallest unit. If there was a IN taking place at a scale smaller than 1ms, we would never catch it with these tools.

A conservative guess for the purposes of this assignment could be 5 ms.

```{r}

rope <- c(-5,0) # we only expect negative effects anyway

BF_1 <- bayesfactor_rope(posterior = xmdl,
                         prior = xmdl_prior,
                         null = rope,
                         parameter = "b_voicingvoiceless")

BF_1
# BF = 1.28 is VERY anectodal evidence against the null.
# If I make the rope (-5,5), then BF = 2.70, which is still quite weak.
```

## (c) Check robustness of your analysis

How robust is your conclusion against different prior choices?

```{r}

# try narrower priors
skinny_priors <- c(prior(normal(150, 10), class = 'Intercept'), # baseline
               prior(normal(0, 10), class = 'b', coef = 'voicingvoiceless'), # the voicing difference
               prior(cauchy(0,100), class = 'sd'), # all sources of variance
               prior(cauchy(0,100), class = 'sigma'), # residual variance
               prior(lkj(1), class = 'cor')) # Don't change this one

# intercept is the mean of 4 random subjects; we assume a less conservative 15 ms IN effect
random_mean <- incNeutr %>% filter(subject %in% sample(unique(subject), 4)) %>% group_by(subject) %>% ungroup() %>% summarize(mean = mean(vowel_dur)) 
random_mean <- 173 # I got 173; run this for repeatability or the line above alone for a random one

semirandom_priors <- c(prior(normal(random_mean,50), class = 'Intercept'), # baseline: 4 random means
               prior(normal(15, 30), class = 'b', coef = 'voicingvoiceless'), # assume IN has a 15ms effect
               prior(cauchy(0,100), class = 'sd'), # all sources of variance
               prior(cauchy(0,100), class = 'sigma'), # residual variance
               prior(lkj(1), class = 'cor')) # Don't change this one

# assume wide, basically flat, priors for intercept and the IN effect
huge_priors <- c(prior(normal(0, 5000), class = 'Intercept'), # baseline
               prior(normal(0, 5000), class = 'b', coef = 'voicingvoiceless'), # the voicing difference
               prior(cauchy(0,100), class = 'sd'), # all sources of variance
               prior(cauchy(0,100), class = 'sigma'), # residual variance
               prior(lkj(1), class = 'cor')) # Don't change this one

# crank up the sd and sigma factors to be much wider to maybe absorb more variance
soaker_priors <- c(prior(normal(150,50), class = 'Intercept'), # baseline
               prior(normal(0, 30), class = 'b', coef = 'voicingvoiceless'), # the voicing difference
               prior(cauchy(0,300), class = 'sd'), # all sources of variance
               prior(cauchy(0,300), class = 'sigma'), # residual variance
               prior(lkj(1), class = 'cor')) # Don't change this one


test_a_prior <- function(new_prior) {
  xmdl_prior <- brm(mdl_formula,
            family = gaussian(),
            prior = new_prior,
            sample_prior = "only",
            data = incNeutr,
            seed = 42)
  
  BF <- bayesfactor_rope(posterior = xmdl,
                         prior = xmdl_prior,
                         null = rope,
                         parameter = "b_voicingvoiceless")
  return(BF)
}

BF_skinny <- test_a_prior(skinny_priors)      # BF = 4.23...over course it went up, but not by much
BF_random <- test_a_prior(semirandom_priors)  # BF = 1.12...about the same
BF_huge <- test_a_prior(huge_priors)          # BF = 0.007...of course it went down
BF_soaker <- test_a_prior(soaker_priors)      # BF = 1.37...about the same

# the conclusion (that there is insufficient evidence to reject the nullish region) holds across all the priors.
```

# References

<!-- References will auto-populate in the refs div below -->

::: {#refs}
:::
