---
title: "session09 - exercises"
format: html
editor: visual
date: 2025-07-17
author: Timo Roettger
execute:
  error: false
  warning: false
  message: false
  cache: false
bibliography: ../resources/bibliography.bib
---

# Preamble: Loading packages and configuration

```{r}
#| label: data_and_libraries
#| echo: false

# just run this code chunk
# function to ignoring the setting of the relative path below when knitting
run_if_not_knitting <- function(expr) {
  if (!isTRUE(getOption("knitr.in.progress"))) {
    eval(expr)
  }
}

# nifty code using the pacman package
# it checks if the packages specified below are installed, if not, they will be installed, if yes, they will be loaded
if (!require("pacman")) install.packages("pacman")
pacman::p_load(rstudioapi, tidyverse, lme4)

# set the current working directory to the one where this file is
run_if_not_knitting(current_working_dir <- dirname(rstudioapi::getActiveDocumentContext()$path))
run_if_not_knitting(setwd(current_working_dir))

```

# Introduction

The main **learning goals** of this week's practical parts are:

-   getting an intuition about the gap between verbal hypotheses and formal hypotheses

# The data

In this dataset, we will be analyzing data from @kilpatrick2023random:

In a nutshell, the authors were interested in whether there is a relationship between the form of a word and its meaning by looking at Japanese Pokemon names.

Let's have a look at the data:

```{r}
#| label: data

# Load in the dataset and already remove irrelevant stuff:
pokemon <- read_csv('../../data/Japanese_Pokemon.csv') |> 
  select(-`Sum of values`, -Checksum)

# check data set:
pokemon

```

What you see here are the following variables:

Variables:

`Designation`: source of Pokemon name. Official = the official names of Japanese Pokemon; Elicited = elicited names from naive participants.

(`pokemonname`: Pokemon name in Japanese) removed in simplified data file

`Output`: Pokemon name in transcribed form

`Length`: Amount of phones in name

`Evolution`: Evolution stage of Pokemons: 1 = pre-evolution; 2 = post-evolution, NA = do not evolve

`f` = amount of /f/ sounds in name

`T` = amount of /T/ sounds in name

...

and many more different sounds.

Now imagine researchers want to use this data set to *test* the following hypothesis:

*The evolution stage of a Pokemon is reflected in aspects of its Japanese name*

# Exercise

## A: Preregister your analysis

Take the hypothesis and what you know about the data set and its structure to predict a very specific relationship that you can test. You are allowed to look at the dataset, but do not look at concrete relationships that might be a manifestation of the hypothesis.

```{r}
# I predict that pokemon that are further along in their evolution will have longer names. 2 is longer than N/A is longer than 1.
pacman::p_load(rstudioapi, tidyverse, tidybayes, brms, bayesplot, bayestestR, modelr)

main_formula <- bf(Length ~ Evolution) # the semantic content predicts the form

get_prior(main_formula, pokemon) # priors are b_Evolution and Intercept

# mean length for all names is ~7.7
mean(pokemon$Length)

# there are 785 1s, 270 NAs, and 810 2s 
table(pokemon$Evolution)
colSums(is.na(pokemon))[5]

# naive priors
my_priors <- c(
  prior(normal(7.7, 4), class = 'Intercept'),
  prior(normal(2, 4), class = 'b')
)


```

## B: Run your preregistered analysis

Run the preregistered analysis made in (A) above and interpret your results relative to the hypothesis.

```{r}

# we'll do a linear model why not?
model <- lm(main_formula, data = pokemon)
coef(model)
# Intercept is 5.7, Evolution is 1.3. Tiny p values (< 2.2e-16 for everything)



model_posterior <- brm(main_formula,
                  family = gaussian(),
                  prior = my_priors,
                  data = pokemon,
                  chains = 4,
                  cores = 4, 
                  iter = 2000,
                  warmup = 1000,
                  thin = 1,
                  seed = 42)

model_posterior # ~4000 Bulk_ESS and ~3000 Tail_ESS for all params--looks good
mcmc_plot(model_posterior, type = 'trace') # good caterpillars
# Intercept is ~5.7 (5.40-5.99) and Evolution is ~1.28 (1.09-1.46). Good confidence!

# It's settled: more evolved pokemon have longer names. And we did it with both Bayes and frequentism, so it should pass everyone's peer reviews :)
```

## C: Show contradicting results

Now explore the data and find two alternative way to test the research hypothesis that yield convincing results that go in opposite directions based of your findings in (A). Justify all your choices verbally after the fact.

```{r}
# soo we want to show that shorter = more evolved

# maybe the final form is more likely to end in /n/ (written as [X] here)
#pokemon_n <- 
#  pokemon %>% mutate(nasal_end = endsWith(Output, 'X'))


# maybe we think, no no no, LENGTH should be the predictor, and evolution should be the DV, since it is probably maybe the length of a proposed name that affects which level of evolution the pokemon designers feel like it should be assigned to. 
# That means we are predicting a categorical variable with a continuous variable. I'm not sure if we learned how to do that with Bayes...I'll just use a glm. 

# We will have to recode 1, 2, NA to a 0-1 scale
pokemon_new_evo <- pokemon %>% 
  mutate(
    new_evo = case_when(
      Evolution == 1 ~ 0,
      Evolution == 2 ~ 1,
      Evolution == NA ~ 0.5 # This includes both single-stage and mid-stage pokemon
    )
  )
contrarian_formula_1 <- bf(new_evo ~ Length)
model_converse <- glm(contrarian_formula_1, 
                      data = pokemon_new_evo, 
                      family = binomial)
# Intercept is -2.7

newdata <- data.frame(Length = seq(1, 20, length.out = 30))
preds <- predict(model_converse, newdata = newdata, type = "response", se.fit = TRUE)
newdata$pred_evo <- preds$fit
ggplot(newdata, aes(x = Length, y = pred_evo)) + geom_point()


# And then maybe some other researchers come along and say, "Oh you silly geese, your analysis doesn't include NAs properly! We need to recode everything."

contrarian_formula_1 = bf(Length ~ Evolution + X + (1 + Evolution | Designation) )
linear_model_n <- lmer(contrarian_formula_1, data = pokemon)
coef(linear_model_n)

# maybe we need some more random effects.

# maybe they end in a vowel?


# And then maybe some other researchers come by and say "Folks I love what you did with the recoding of the evolutionary scale. But GLMs have p-values and Timo told me not to touch that. If it's not Bayes, there's no ways I use it. We'll replicate your analysis with Bayesian methods--but we also think that we need to account for which sound the name ends with.

```

## D: Researcher degrees of freedom

Note down all dimensions of researcher degrees of freedom that make the mapping of the hypothesis onto your data analysis vague. How many possible combinations can you think of?

# References

<!-- References will auto-populate in the refs div below -->

::: {#refs}
:::
