---
title: "TORRUELLA_07_homework"
format: html
---

---
title: "session07 - homework"
format: html
editor: visual
date: 2025-07-17
author: Timo Roettger
execute:
  error: false
  warning: false
  message: false
  cache: false
---

# Preamble: Loading packages and configuration

```{r}
#| label: data_and_libraries
#| echo: false

# just run this code chunk
# function to ignoring the setting of the relative path below when knitting
run_if_not_knitting <- function(expr) {
  if (!isTRUE(getOption("knitr.in.progress"))) {
    eval(expr)
  }
}

# nifty code using the pacman package
# it checks if the packages specified below are installed, if not, they will be installed, if yes, they will be loaded
if (!require("pacman")) install.packages("pacman")
pacman::p_load(rstudioapi, tidyverse, lme4, brms, bayestestR)

# set the current working directory to the one where this file is
run_if_not_knitting(current_working_dir <- dirname(rstudioapi::getActiveDocumentContext()$path))
#run_if_not_knitting(setwd(current_working_dir))

# options to increase efficiency of brms models (optional)
#options(brms.backend = "cmdstanr")
#options(mc.cores = parallel::detectCores())

```

### Introduction to the problem

We will be working with Quantese again, see `01_exercise.qmd` and the other data resources for a refresher.

```{r}
#| label: load-quantese

quantese <- read_csv("../data/quantese_naming_sim.csv") 
# add log-transformed reaction time
quantese <- quantese |> 
  mutate(log_rt = log(rt))

# have a look
quantese

```

In an earlier homework, we have assessed the following hypothesis:

*"We hypothesize that log-transformed reaction times are affected by lexical frequency"*

We tried to run the following model that included an appropriate random effect structure but it did not converge with `lme4`:

```{r}
#| label: lmer_quantese

# formulate the corresponding LMER
quant_lmer <- lmer(log_rt ~ log_frequency +
                     (1 + log_frequency | PartID) +
                     (1 | ItemID), 
                    data = quantese)
# your receive a warning: "boundary (singular) fit: see help('isSingular')"

```

Now we will re-run this model within the Bayesian framework:

### (A) Specify priors

Check which priors this model needs and specify weakly informative priors.

Remember that weakly informative priors are priors that do not bias the posteriors in any direction, but reasonably constrain the parameter space, i.e. only reasonable values are assumed.

For priors of class `Intercept` and `b`, constrain the parameter space with a wide normal distribution.

```{r}
#| label: get_prior

da_formuler = bf(log_rt ~ log_frequency +
                     (1 + log_frequency | PartID) +
                     (1 | ItemID))

# what priors do we need
get_prior(da_formuler, quantese) 
# Intercept (ie distribution of possible average log RTs)
# b:log_frequency (ie distribution of possible differences each unit of log freq has on log RT)

# Intercept
range(quantese$log_rt) # logged reaction times range from ~5.9 to ~7.8. Assuming a normal distribution, 6.85 is the mean. Assuming 2 SDs should capture most of the data, we could do an SD of 0.9.

ggplot(data = quantese) +
  geom_density(aes(x = log_rt, fill = PartID, alpha = 0.5)) +
  stat_function(fun = dnorm, args = list(mean = 6.85, sd = 0.9)) +
  scale_x_continuous(limits = c(0,10))

# This distribution looks reasonably conservative and naive, while still being appropriate to what we know about the nature of the task. Is there any other way of producing weakly informative priors besides taking the range of the data and loosely basing it around that?? Since that is still some sort of posterior analysis feeding back, and not totally naive...maybe if I knew the literature more surrounding this type of study, I could start from a mean/range of log_rts from similar tasks in the past. 
# FYI the conservative student_t recommended is also quite close to this, I think. At least the mean looks to be 6.5.

# log_frequency
ggplot(data = quantese) +
  geom_boxplot(aes(group = log_frequency , y = log_rt, fill = log_frequency))
# Each group usually spans from about 6 to 8. 
# We want to assume naive prior knowledge, meaning we aren't aware of an effect, right? So we have a prior effect size of mean 0. The "ambient jitter" (SD?) of this data is about 1.5 within each log_frequency value, so, if we were assuming that all log_frequencies are equal, that would be a fair SD to give.

# specify weakly informative priors
# To be extra generous in the weakness of our priors, I'll let the SDs be a bit bigger (after all we're going after a "wide normal distribution"), while still be reasonable for the domain (i.e. no near-instantaneous RTs)
my_priors <- c(prior(normal(6.85, 1.5), class = 'Intercept'),
               prior(normal(0, 2), class = 'b', coef = 'log_frequency'))

```

We can also now get used to specifying priors for classes `sd` and `cor`. Priors of class `sd` are priors for the variance associated with our random effects. You see here 2 `sd` priors for group `ItemID`. The first one is just a generic prior for all `ItemID` coefficients and the second one with the `coef = Intercept` stands for the random intercept variation between items. Similarly for group `PartID`, we have 3 priors. The first one is again just a generic prior for all `PartID` coefficients. The second one with the `coef = Intercept` stands for the random intercept variation between participants. And the last one with the `coef = log_frequency` stands for the random slope variation between participants for the predictor `log_frequency`. How do we specify these priors. We have usually no idea what to expect from these parameters, so we here also specify very diffuse priors centered on zero. We pick a cauchy distribution (remember? super squashed normal distributions) because they have thick tails. Let's make them super diffuse and pick a `sd = 3` for both items and participants.

```{r}
#| label: add_sd_priors

# add weakly informative priors for sd
my_priors <- c(my_priors, 
               prior(cauchy(0, 3), class = 'sd', group = 'ItemID'),
               prior(cauchy(0, 3), class = 'sd', group = 'PartID'))

```

And finally, if your model as random slopes, they have by default a correlation parameter for intercept and slope. We are not getting into the math here and will follow advice from smart people. We use Lewandowski-Kurowicka-Joe priors (shape = 1): `lkj(1)`. Specifying this prior for random slope models actually helps a lot with the fitting procedure.

```{r}
#| label: add_cor_priors

# add weakly informative priors for cor
my_priors <- c(my_priors, 
               prior(lkj(1), class = 'cor'))

```

### (B) Run model

Now run the model with 4 chains, 1000 warm-up and 1000 post-warm-up samples per chain. Check that chains are reasonably mixed and that sampling yielded a sufficient effective sample size. If not address the issues appropriately (check slides again).

```{r}
#| label: model_quant

# if the model takes way too long to sample (more than 5-10 mins), use the following subset
# quantese_sub <- quantese |> 
#   filter(PartID %in% c("01","02","03","04",
#                        "05","06","07","08"))

quantese_model <- brm(da_formuler,
            family = gaussian(),
            prior = my_priors,
            data = quantese,
            chains = 4,
            cores = 4, 
            iter = 2000,
            warmup = 1000,
            thin = 1,
            seed = 42)
# Output:
# Warning: Bulk Effective Samples Size (ESS) is too low, indicating posterior means and medians may be unreliable.
# Running the chains for more iterations may help. See
# https://mc-stan.org/misc/warnings.html#bulk-ess
# Warning: Tail Effective Samples Size (ESS) is too low, indicating posterior variances and tail quantiles may be unreliable.
# Running the chains for more iterations may help. See
# https://mc-stan.org/misc/warnings.html#tail-ess

mcmc_plot(quantese_model, type = 'trace') # decent caterpillars
```

### (C) Plot posterior predictions

Find a way to extract posterior samples from the model for the `Intercept` and the `log_frequency` coefficient and plot the predicted posterior mean linear model alongside its 95% Credible Intervals. There are many different ways to get there.

```{r}
#| label: plot_predictions

post_samps <- posterior_samples(quantese_model) # not sure I understand all the stuff (coefficients) this is outputting?

# this just hardcodes a line (not a model) but without confidence intervals or anything :(
mean_intercept <- mean(post_samps$Intercept)
mean_slope <- mean(post_samps$b_log_frequency)
ggplot() + 
  geom_abline(intercept = mean_intercept, slope = mean_slope, color = "black") +
  xlim(0, 10) + ylim(5, 8)

# uhh does this do it? Or does it just show that as a model's intercept is higher, it predicts a stronger negative effect?
ggplot(data = post_samps, aes(x = Intercept, y = b_log_frequency)) +
  geom_smooth(method = 'lm') +
  geom_point(alpha=0.05)


```
